---
title: "Taylor series: basics and visuals"
author: "Lluc Sementé"
date: "9/11/2018"
output: pdf_document

---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(scales)
library(gridExtra)
```

## 1.Taylor series

A one-dimensional **Taylor series**[1] is an expansion of a real function $f(x)$ about a point $x=a$ and it's given by:

$$ f(x) = f(a)+\frac{f'(a)}{1!}(x-a)+\frac{f''(a)}{2!}(x-a)^2+\frac{f^{(3)}(a)}{3!}(x-a)^3+\dots = \sum_{n=0}^{\infty}\frac{f^{(n)}(a)}{n!}(x-a)^n $$
This sereies expansion transforms information about the derivative terms of the function into local information around a point. As we keep adding terms in the series, the resulting polynomial will fit better and better the function in the convergence interval. If a = 0, the expansion is known as a **Maclaurin series**.

Taylor series are useful to simplify complex functions around a point in order to ease the calculus in which are involved. This is a specially crucial step in some common physics problems. Probably, the most used Taylor series approximation is $cos(\theta) = 1 - \frac{\theta^2}{2}$, commonly used in pendulum oscillation related problems.

Let's now end with boring formulas and enter the beautiful world of visualization.

## 2.Maclaurin series of common functions

Time to focus our attention into some well known functions and its Maclaurin series through intuitive colorful plots. Let's start with the trigonometric functions cosine and sine.

### 2.1 Trigonometric related functions

First, let's look at the functions alone. 

```{r, echo=FALSE, message=FALSE, warning=FALSE,dpi=300,fig.align='center',fig.width=7,fig.height=3.5}
x <- seq(from = -2*pi, to = 2*pi,length.out = 32768)
y <- cos(x)
cosine <- data.frame("x" = x, "y" = y)

x <- seq(from = -2*pi, to = 2*pi,length.out = 32768)
y <- sin(x)
sine <- data.frame("x" = x, "y" = y)
pi_scales <- math_format(.x * pi, format = function(x) x / pi)

p1 <- ggplot(data = cosine, mapping = aes(x = x, y = y)) +  geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-2*pi,0,2*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(colour="red") + xlim(c(-2*pi,2*pi)) + ylim(c(-2,2)) + theme_bw() + scale_x_continuous(labels = pi_scales, breaks = seq(from = -2*pi,  to = 2*pi, by = (pi/2)),expand = c(0,0)) + xlab("x") + ylab("y") + scale_y_continuous(breaks = c(-1,0,1),limits = c(-2,2)) + ggtitle("Cos(x)")

p2 <- ggplot(data = sine, mapping = aes(x = x, y = y)) +  geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-2*pi,0,2*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(colour="red") + xlim(c(-2*pi,2*pi)) + ylim(c(-2,2)) + theme_bw() + scale_x_continuous(labels = pi_scales, breaks = seq(from = -2*pi,  to = 2*pi, by = (pi/2)),expand = c(0,0)) + xlab("x") + ylab("y") + scale_y_continuous(breaks = c(-1,0,1),limits = c(-2,2)) + ggtitle("Sin(x)")

grid.arrange(p1,p2,nrow = 1)
```

Applying the definition of Taylor series, the cosine's and sine's Taylor series around $x=0$ (Maclaurin series) looks like this:
$$ cos(x) = 1-\frac{x^2}{2!}+\frac{x^4}{4!}-\frac{x^6}{6!}+\dots = \sum_{n=0}^{\infty}(-1)^{n}\frac{x^{2n}}{2n!} $$
$$sin(x) = {x}-\frac{x^3}{3!}+\frac{x^5}{5!}-\frac{x^7}{7!}+\dots = \sum_{n=0}^{\infty}(-1)^{n}\frac{x^{2n+1}}{(2n+1)!}$$


Using this formulas, let's see how good is this tool to approximate the behavior of $cos(x)$ and $sin(x)$ around $x=0$.
```{r, echo=FALSE, message=FALSE, warning=FALSE,dpi=300,fig.align='center',fig.width=9.92,fig.height=3.5}
x <- seq(from = -2*pi, to = 2*pi,length.out = 32768)
y <- c((1-((x^2)/2)),
       (1-((x^2)/2)+((x^4)/factorial(4))),
       (1-((x^2)/2)+((x^4)/factorial(4))-((x^6)/factorial(6))),
       (1-((x^2)/2)+((x^4)/factorial(4))-((x^6)/factorial(6))+((x^8)/factorial(8))),
       cos(x)
       )
colour <- as.factor(c(rep("2nd degree",times=32768), rep("4th degree",times=32768), rep("6th degree",times=32768), rep("8th degree",times=32768),rep("cos(x)",times=32768)))

linetype <- as.factor(c(rep(2, times = (4*32768)),rep(1, times = 32768)))
df <- data.frame("x" = x, "y" = y, "colour" = colour, "linetype" = linetype)
pi_scales <- math_format(.x * pi, format = function(x) x / pi)

p1 <- ggplot(data = df, mapping = aes(x = x, y = y)) +  
geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-2*pi,0,2*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(mapping = aes(linetype = linetype, colour = colour)) + 
xlim(c(-2*pi,2*pi)) + ylim(c(-2,2)) + 
theme_bw() + 
scale_x_continuous(labels = pi_scales, breaks = seq(from = -2*pi,  to = 2*pi, by = (pi/2)),expand = c(0,0)) + 
xlab("x") + ylab("y") + 
scale_y_continuous(breaks = c(-1,0,1),limits = c(-2,2)) +
scale_linetype_manual(values = c("solid", "dashed"), guide = FALSE) + ggtitle("Cosine's Maclaurin series") + labs("Function") +
scale_color_manual(values = c("cos(x)" = "red","2nd degree" = "#F8766D", "4th degree" = "#00BA38", "6th degree" = "#00B0F6", "8th degree" ="#FD61D1" ) )

x <- seq(from = -2*pi, to = 2*pi,length.out = 32768)
y <- c(sin(x),
       (x),
       (x)-((x^3)/factorial(3)),
       (x)-((x^3)/factorial(3))+((x^5)/factorial(5)),
       ((x)-((x^3)/factorial(3))+((x^5)/factorial(5))-((x^7)/factorial(7)))
       )
colour <- as.factor(c(rep("sin(x)",times=32768), rep("1st degree",times=32768), rep("3rd degree",times=32768), rep("5th degree",times=32768), rep("7th degree",times=32768)))

linetype <- as.factor(c(rep(1, times = 32768), rep(2, times = (4*32768))))
df <- data.frame("x" = x, "y" = y, "colour" = colour, "linetype" = linetype)
pi_scales <- math_format(.x * pi, format = function(x) x / pi)

p2 <- ggplot(data = df, mapping = aes(x = x, y = y)) +  
geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-2*pi,0,2*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(mapping = aes(linetype = linetype, colour = colour)) + 
xlim(c(-2*pi,2*pi)) + ylim(c(-2,2)) + 
theme_bw() + 
scale_x_continuous(labels = pi_scales, breaks = seq(from = -2*pi,  to = 2*pi, by = (pi/2)),expand = c(0,0)) + 
xlab("x") + ylab("y") + 
scale_y_continuous(breaks = c(-1,0,1),limits = c(-2,2)) +
scale_size_manual(values = c(1, 0.8), guide = FALSE) +
scale_linetype_manual(values = c("solid", "longdash"), guide = FALSE) + 
ggtitle("Sine's Maclaurin series") + labs("Function") +
scale_color_manual(values = c("sin(x)" = "red","1st degree" = "#F8766D", "3rd degree" = "#00BA38", "5th degree" = "#00B0F6", "7th degree"="#FD61D1" ) )
        

grid.arrange(p1,p2,nrow = 1)
```

That looks smooth! Notice that, as we keep adding polynomial terms, the approximation is getting better and better around $x=0$.
Also, cosine is an even function ($i.e. cos(-x) = cos(x)$) and his Maclaurin series has only even powers, and sine is an odd function ($i.e. sin(-x) = -sin(x)$) and his Maclaurin series has only odd powers. What kind of black magic is this?
Also it's beautiful to see the patterns that are formed over-plotting each consecutive polynomial and how the inner symmetry of each function is visualized.

Time for the cardinal version of the sine function, a very important function in signal processing. Trying to obtain the derivatives of the cardinal sine can be overwelming, so instead, let's use the sine's Maclaurin serie:

$$ sinc(x) = \frac{sin(x)}{x} = \frac{\sum_{n=0}^{\infty}(-1)^{n}\frac{x^{2n+1}}{(2n+1)!}}{x} = \sum_{n=0}^{\infty}(-1)^{n}\frac{x^{2n}}{(2n+1)!}$$
```{r, echo=FALSE, message=FALSE, warning=FALSE,dpi=300,fig.align='center',fig.width=9.92,fig.height=3.5}

x <- seq(from = -4*pi, to = 4*pi,length.out = 3*32768)
y <- c(sin(x)/x)
df <- data.frame("x" = x, "y" = y)

pi_scales <- math_format(.x * pi, format = function(x) x / pi)

p1 <- ggplot(data = df, mapping = aes(x = x, y = y)) +  geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-4*pi,0,4*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(colour="red") + xlim(c(-4*pi,4*pi)) + ylim(c(-0.5,1.5)) + theme_bw() + scale_x_continuous(labels = pi_scales, breaks = seq(from = -4*pi,  to = 4*pi, by = (pi)),expand = c(0,0)) + xlab("x") + ylab("y") + scale_y_continuous(breaks = c(0,1),limits = c(-0.5,1.5)) + ggtitle("Sinc(x)")


x <- seq(from = -4*pi, to = 4*pi,length.out = 3*32768)
y <- c(sin(x)/x,
       ((1)-(x^2)/factorial(3)),
       ((1)-((x^2)/factorial(3))+((x^4)/factorial(5))),
       ((1)-((x^2)/factorial(3))+((x^4)/factorial(5))-((x^6)/factorial(7))),
       ((1)-((x^2)/factorial(3))+((x^4)/factorial(5))-((x^6)/factorial(7))+((x^8)/factorial(9)))
       )
colour <- as.factor(c(rep("sinc(x)",times=3*32768), rep("2nd degree",times=3*32768), rep("4th degree",times=3*32768), rep("6th degree",times=3*32768), rep("8th degree",times=3*32768)))

linetype <- as.factor(c(rep(1, times = 3*32768), rep(2, times = (4*3*32768))))
df <- data.frame("x" = x, "y" = y, "colour" = colour, "linetype" = linetype)
pi_scales <- math_format(.x * pi, format = function(x) x / pi)

p2 <- ggplot(data = df, mapping = aes(x = x, y = y)) +  
geom_hline(yintercept = c(-1,0,1),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_vline(xintercept = c(-4*pi,0,4*pi),linetype = c("dashed","longdash","dashed"),colour = c("grey","black","grey")) + geom_line(mapping = aes(linetype = linetype, colour = colour)) + 
xlim(c(-4*pi,4*pi)) + ylim(c(-2,2)) + 
theme_bw() + 
scale_x_continuous(labels = pi_scales, breaks = seq(from = -4*pi,  to = 4*pi, by = (pi)),expand = c(0,0)) + 
xlab("x") + ylab("y") + 
scale_y_continuous(breaks = c(0,1),limits = c(-0.5,1.5)) +
scale_size_manual(values = c(1, 0.8), guide = FALSE) +
scale_linetype_manual(values = c("solid", "longdash"), guide = FALSE) + 
ggtitle("Cardinal sine's Maclaurin series") + labs("Function") +
scale_color_manual(values = c("sinc(x)" = "red","2nd degree" = "#F8766D", "4th degree" = "#00BA38", "6th degree" = "#00B0F6", "8th degree"="#FD61D1" ) )

grid.arrange(p1,p2,nrow = 1)
```

### 2.1 Exponential related functions

Time for some exponential related functions like the natural exponential and the Gaussian function:

```{r, echo=FALSE, message=FALSE, warning=FALSE,dpi=300,fig.align='center',fig.width=7,fig.height=3.5}
x <- seq(from = -5, to = 5,length.out = 32768)
y <- exp(x)
p.exponential <- data.frame("x" = x, "y" = y)

p1 <- ggplot(data = p.exponential, mapping = aes(x = x, y = y)) +
geom_line(colour="red") + 
xlim(c(-5,5)) + 
ylim(c(-1,9)) + 
theme_bw() + 
scale_x_continuous(breaks = seq(from = -5, to = 5, by = 1),expand = c(0,0)) + 
xlab("x") + 
ylab("y") + 
scale_y_continuous(breaks = seq(from = -1, to = 9, by = 1),limits = c(-1,9),expand = c(0,0)) +
ggtitle("Positive natural exponential")


x <- seq(from = -5, to = 5,length.out = 32768)
y <- exp(-(x^2))
n.exponential <- data.frame("x" = x, "y" = y)

p2 <- ggplot(data = n.exponential, mapping = aes(x = x, y = y)) +
geom_line(colour="red") + 
theme_bw() + 
scale_x_continuous(breaks = seq(from = -3, to = 3, by = 1),limits = c(-3,3),expand = c(0,0)) + 
xlab("x") + 
ylab("y") + 
scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.25),limits = c(-0.25,1.25),expand = c(0,0)) +
ggtitle("Gaussian function")


grid.arrange(p1,p2,nrow = 1)
```
In this case, the Taylor's serie at $x=e$ is going to be displayed for the positive natural exponential, and the Maclaurin for the Gaussian function. But first, the formulas:

```{r, echo=FALSE, message=FALSE, warning=FALSE,dpi=300,fig.align='center',fig.width=7,fig.height=3.5}
x <- rep(seq(from = -5, to = 5,length.out = 32768), times = 2)
y <- exp(x)
p.exponential <- data.frame("x" = x, "y" = y)

p1 <- ggplot(data = p.exponential, mapping = aes(x = x, y = y)) +
geom_line(colour="red") + 
xlim(c(-5,5)) + 
ylim(c(-1,9)) + 
theme_bw() + 
scale_x_continuous(breaks = seq(from = -5, to = 5, by = 1),expand = c(0,0)) + 
xlab("x") + 
ylab("y") + 
scale_y_continuous(breaks = seq(from = -1, to = 9, by = 1),limits = c(-1,9),expand = c(0,0)) +
ggtitle("Positive natural exponential")


x <- seq(from = -5, to = 5,length.out = 32768)
y <- exp(-(x^2))
n.exponential <- data.frame("x" = x, "y" = y)

p2 <- ggplot(data = n.exponential, mapping = aes(x = x, y = y)) +
geom_line(colour="red") + 
theme_bw() + 
scale_x_continuous(breaks = seq(from = -3, to = 3, by = 1),limits = c(-3,3),expand = c(0,0)) + 
xlab("x") + 
ylab("y") + 
scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.25),limits = c(-0.25,1.25),expand = c(0,0)) +
ggtitle("Gaussian function")


grid.arrange(p1,p2,nrow = 1)
```


## References

[1] Weisstein, Eric W. "Taylor Series." From MathWorld--A Wolfram Web Resource. http://mathworld.wolfram.com/TaylorSeries.html